---
title: "Iteration With purrr"
author: "JMG"
date: "11/3/2021"
output: powerpoint_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(patchwork)
library(latex2exp)
library(kableExtra)
```

## Learning Objectives

In this lesson, we cover the basic usage and application of iteration following "tidy" principles by using the `purrr` package. After this lesson you should

- Be familiar with common loop patterns.
- Have a feel for the map family of `purrr` functions. 
- Be aware of some statistical applications of iteration using the `purrr` package.


## References 

References for this lesson are

- [R for Data Science](https://r4ds.had.co.nz/)
- [Advanced R](https://adv-r.hadley.nz/)
- [purrr Documentation](https://purrr.tidyverse.org/)

You can find code relevant to this lesson on [my GitHub repo](https://github.com/jmgraham30/SWBmodules). 

## Prerequisites

This lesson assumes familiarity with

- The use of [tidyverse](https://www.tidyverse.org/) packages such as `tibble`, `ggplot2`, and `dplyr`. 
  - In particular, we will assume that the learner is comfortable with working with data frames or tibbles and is familiar with basic plotting using `ggplot2`.  

- We will refer to some basic statistical concepts such as confidence intervals and linear models. However, you can probably still get a fair amount out of this lesson even if your statistics background is minimal.  

## Iteration 


## Functional Programming




## R Markdown

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Slide with Bullets

- Bullet 1
- Bullet 2
- Bullet 3

## Slide with R Output

```{r cars, echo = TRUE}
summary(cars)
```

## Slide with Plot

```{r pressure}
plot(pressure)
```

