---
title: "Dates and Times in R"
author: "Jason M Graham"
date: "`r Sys.Date()`"
output: 
  powerpoint_presentation:
    reference_doc: BasicStatsRTemplate.pptx
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(lubridate)
library(nycflights13)
library(flextable)
# https://datausa.io/profile/naics/restaurants-food-services
data_folder <- "../Data/"
data_file <- paste0(data_folder,"Monthly Employment.csv")
food_services <- read_csv(data_file)
r_package_downloads <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2018/2018-10-30/r-downloads.csv")
```

## Goals

After this lesson, you should

1) be aware of some common challenges of working with data that include dates and times, and

2) know some introductory level techniques for handling date-time data in R using the [`lubridate`](https://lubridate.tidyverse.org/) package.

## Overview

In this lesson, we will

- motivate the need to work with data sets that include variables that record information corresponding to dates or times, 

- explain some common problems associated with data-time data, 

- explain the most basic functions from the `lubridate` package, and

- present some use case examples. 

## Background and Motivation 

- Data comes in a variety of types.

- We measure quantities numerically with decimal numbers. In R, represented by `numeric`.

- Count certain observations with whole numbers. In R, represented by `numeric`.

- Record qualitative or categorical observations using text-based labeling. In R, represented by `character`.

## R Data Examples

```{r,echo=TRUE}
numeric_values <- c(1.2,6.5,8.1)
class(numeric_values)
```

```{r,echo=TRUE}
character_values <- c("dog","cat","bird")
class(character_values)
```

## Data Frames and Examples

- Often, we store a data set as a `data.frame` or `tibble` in R.

  - Each column in a data frame can have a different type for the data it contains.
  
- We will work with three data sets:

1) The restaurants and food service data from [Data USA](https://datausa.io/) contained in the `Monthly Employment.csv` file,

2) the [R and R Packages download data](https://github.com/rfordatascience/tidytuesday/tree/master/data/2018/2018-10-30) from [Tidy Tuesday](https://github.com/rfordatascience/tidytuesday), and

3) The airline flights data from the `nycflights13` package. 

## Looking at the Data

- Let's look at what is in the data frames for our working examples data.

  - We will show the first few rows for the food services and R package downloads data sets. 
  
## Food Services

```{r}
head(food_services,4) %>% flextable()
```

- What do you think are the data types for each column of the data? 

## R Package Downloads

```{r}
head(r_package_downloads,4) %>% flextable()
```

- What do you think are the data types for each column of the data?

## Date and Time Data

- Some variables in the data sets have a `date` or `time` type. 

- We can see the types for the columns in the data frames with,

```
glimpse(food_services)
glimpse(r_package_downloads)
```

- The output shows that the date and time columns have data types associated with a `date` or `time` type in R.

## Date-Time Challenges

Some common challenges of working with data that contains date-time entries include 

- Different individuals record dates and times differently. 

- For example, all of the following represent the same day:

  * February 14, 2022

  * 02-14-2022

  * 2022-02-14

  * 14-Feb-2022


## Date and Time Challenges in a Spreadsheet

- It can be very challenging to work with date or time data in a spreadsheet.

- Working in R facilitates easier handling of date-time data, especially when using the `lubridate` package.  

## The `lubridate` Package

- [`lubridate`](https://lubridate.tidyverse.org/) is an R package associated with the `tidyverse` family of packages

- `library(lubridate)` loads the package, note that `lubridate` is not loaded by `library(tidyverse)`

- `lubridate` usage in covered in the [Dates and Times chapter](https://r4ds.had.co.nz/dates-and-times.html) of [R for Data Science](https://r4ds.had.co.nz/index.html). 

- Let's see some of the functionality of `lubridate`. 

## Creating Dates

- `lubridate` contains functions `ymd`, `ydm`, `mdy`, `myd`, `dmy`, and `dym` to parse a string or character into a date. For example,

```{r,echo=TRUE}
(ymd("2022-03-06"))
(mdy("March 6th, 2022"))
(dmy("06-March-2022"))
(ymd(20220306))
```


## Creating Date-Times

- We can also parse date-time strings. For example, 

```{r,echo=TRUE}
(ymd_hms("2022-03-06 08:58:31"))
(mdy_hm("03/06/2022 08:58"))
```

## Manipulating Date-Times

- It is also possible to manipulate dates and times with lubridate functions. 

- For example, the following command will split columns for the year, month, and day from the `Date` column

```
food_services %>% 
  mutate(year=year(Date),month=month(Date),day=day(Date)) %>%
  glimpse()
```

- Note that we have used the `lubridate` functions `year`, `month`, and `day` in combination with the `mutate` function from `dplyr` to add three new columns.

## Combining Date Components

- Perhaps a data set has the year, month, and day spread across multiple columns and we would like to combine these into a single column. 

- This is done with a command such as 

```
make_date(year=2022,month=03,day=10)
```

- Let's see an application of this to a data frame. 

## Use Case Examples

- Let's go to RStudio and work through some code for some common use case examples together. 


## Summary

In this lesson we

- motivated the necessity of working with date and time data,

- explained some of the challenges that this present, and

- showed `lubridate` functions that address facilitate solving common date-time problems in R.


## Conclusion

