---
title: "Iteration With purrr"
author: "JMG"
date: "11/3/2021"
output: powerpoint_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(patchwork)
library(latex2exp)
library(kableExtra)
```

## Learning Objectives

In this lesson, we cover the basic usage and application of "tidy" iteration using the `purrr` package. After this lesson you should:

- Be familiar with common loop patterns.
- Have a feel for the map family of `purrr` functions. 
- Be aware of some statistical applications of iteration using the `purrr` package.


## References 

References for this lesson are

- [R for Data Science](https://r4ds.had.co.nz/)
- [Advanced R](https://adv-r.hadley.nz/)
- [purrr Documentation](https://purrr.tidyverse.org/)

You can find code relevant to this lesson on [my GitHub repo](https://github.com/jmgraham30/SWBmodules). 

## Prerequisites

This lesson assumes familiarity with

- Vectors, lists, data frames, and functions in R. 
- The use of [tidyverse](https://www.tidyverse.org/) packages such as [`tibble`](https://tibble.tidyverse.org/), [`ggplot2`](https://ggplot2.tidyverse.org/), and [`dplyr`](https://dplyr.tidyverse.org/). 
 
- Knowledge of basic statistical concepts such as confidence intervals and linear models is helpful but not essential.   

## Iteration 

In [R for Data Science](https://r4ds.had.co.nz/), good programming practices are stressed. For example, it is emphasized that one should never copy and paste code more than twice. 

- Loops provide one way to avoiding copying and pasting, 

- functions provide another.  

Let's review the basic programming construct known as for loops. 


## Loops 


## Loops in R

A `for` loop in R has the general form

`for (item in vector) perform_action`

For example:

```
for (i in 1:3){
   print(2*i)
}
```

## Output From Example Loop

The `for` loop in the previous slide will output
```{r}
for (i in 1:3){
   print(2*i)
}
```


## Functional Programming






