---
title: "Iteration With purrr"
author: "JMG"
date: "11/3/2021"
output: 
  powerpoint_presentation:
    reference_doc: BasicStatsRTemplate.pptx
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(broom)
library(faraway)
library(patchwork)
```

## Learning Objectives

In this lesson, we cover the basic usage and application of "tidy" iteration using the `purrr` package. After this lesson you should:

1. Have a feel for the map family of `purrr` functions. 

2. Have an understanding of how `purr` functions can be used in place of loops.

3. Be aware of some statistical applications of iteration using the `purrr` package.


## References 

References for this lesson are

* [R for Data Science](https://r4ds.had.co.nz/)

* [Advanced R](https://adv-r.hadley.nz/)

* [purrr Documentation](https://purrr.tidyverse.org/)

You can find code relevant to this lesson on [my GitHub repo](https://github.com/jmgraham30/SWBmodules). 

## Prerequisites

This lesson assumes familiarity with

- Vectors, lists, data frames, and functions in R. 

- The use of [tidyverse](https://www.tidyverse.org/) packages such as [`tibble`](https://tibble.tidyverse.org/), [`ggplot2`](https://ggplot2.tidyverse.org/), and [`dplyr`](https://dplyr.tidyverse.org/). 
 
- Knowledge of basic statistical concepts such as confidence intervals and linear models is helpful but not essential.   

## Iteration 

In [R for Data Science](https://r4ds.had.co.nz/), good programming practices are stressed. For example, it is emphasized that one should never copy and paste code more than twice. 

- Loops provide one way to avoiding copying and pasting. 

- Functions provide another.  

Let's review the basic programming construct known as for loops. 


## Loops 

Loops are examples of control flow statements in a program that result in iteration. 

- The basic idea of a for loop is that a specified action is repeated a predetermined number of times. 

- For example, if we have a container that holds three numbers and we want to multiply each of the three numbers by 2 then a for loop is an ideal method for doing this repetitive task.  

## Loops in R

A `for` loop in R has the general form

`for (item in vector) perform_action`

For example:

```
for (i in 1:3){
   print(2*i)
}
```

## Output From Example Loop

The `for` loop in the previous slide will output
```{r}
for (i in 1:3){
   print(2*i)
}
```

Note that if we wanted to store the result of multiplying by 2 instead of printing it, we would need to initialize a vector of the appropriate size. 


## Functional Programming

- Often, for loops are unnecessary in R because the language supports [functional programming](https://en.wikipedia.org/wiki/Functional_programming) (FP).

- FP is a paradigm that utilizes functions and function composition to develop code.

- Use of FP can lead to clean, concise code. 

- The `purrr` package enhances FP in R.  

- FP code is convenient for multi-processing. 

## FP: Getting Started

A good way to get started with FP in R is via the `map` family of functions from `purrr`. 

- Here is a simple example:

```
map(1:3,function(x){2*x})
```

- This creates a list object that contains 2, 4, and 6.

## Output from Previous Slide

```{r}
map(1:3,function(x){2*x})
```

## Explanation of map

- The `map` function inputs an object such as a vector, list, or data frame together with a function, and applies that function to each component of the object. 



## Input/Output for map 

- `map` returns a list object. 

- There are similar functions, `map_lgl`, `map_chr`, `map_dbl`, `map_int`, etc. that return different data types. Examine documentation with `?map` for details. 

- Let's see some examples for `map`. 

- We start by using `map` to solve the same problem we already solved using a loop, that is, to multiply a sequence of numbers by 2.  


## Example Output

- For example, 

```
map_dbl(1:3,function(x){2*x})
```

outputs

```{r}
map_dbl(1:3,function(x){2*x})
```

- This is a simple (atomic) vector instead of a list. 

- Let's see some more interesting use cases. 

## Compute Multi-Column Medians

```{r sim_data,echo=FALSE}
sim_data <- tibble(a=rnorm(57),b=runif(57),c=rpois(57,0.7),d=runif(57,1,3))
```

- Suppose we have a data frame named `sim_data` with 4 numerical columns and we want to compute the median for each column.  

```
map_dbl(sim_data,median)
```
```{r}
round(map_dbl(sim_data,median),2)
```



## Bootstrap Confidence Intervals

- Bootstrapping approximates a sampling distribution by sampling **with replacement** many times the original sample. 

- We will see an application of using `purrr` to implement a bootstrap to approximate a confidence interval for a mean and confidence intervals for parameters in a linear model. 

## Bootstrapping a Mean

- We start with some simulated data:
```
set.seed(42)
x <- rnorm(35,10,6.7)
```

- Then we define a function that will return a single sample with replacement of out data:
```
my_boot_samp <- function(i){
  return(sample(x,replace=TRUE))
}
```

```{r,echo=FALSE}
set.seed(42)
x <- rnorm(35,10,6.7)
my_boot_samp <- function(i){
  return(sample(x,replace=TRUE))
}
```

## Applying map to Obtain Resamples

- Next, we apply `map` to obtain a list of 500 resamples:

```
boot_samps <- map(1:500,my_boot_samp)
```

- Finally, we use `map_dbl` to compute the mean for each of our bootstrap resamples:
```
boot_means <- map_dbl(boot_samps,mean)
```

```{r,echo=FALSE}
boot_samps <- map(1:500,my_boot_samp) 
boot_means <- map_dbl(boot_samps,mean)
```

## Results

- We compare the bootstrap CI with the usual formula:

  - Bootstrap CI
  ```
  quantile(boot_means,probs = c(0.025,0.975))
  ```
  ```{r,echo=FALSE}
  quantile(boot_means,probs = c(0.025,0.975))
  ```
  
  - Usual CI
  ```
  mean(x) + 1.96*c(-1,1)*(sd(x)/sqrt(length(x)))
  ```
  ```{r,echo=FALSE}
  mean(x) + 1.96*c(-1,1)*(sd(x)/sqrt(length(x)))
  ```

## Bootstrap Linear Model Parameters

- Given a function `boot_samp_leafburn` that samples from the dataset, the following code obtains 500 bootstrap resamples of the `leafburn` data from the `faraway` package:
```
leafburn_boots <- map(1:500,boot_samp_leafburn)
```

## Bootstrap Linear Model Parameters, Cont.

- Now, given a function `leafburn_boot_fit` that returns the parameters from a linear fit, we obtain bootstrapped estimates:

```
boot_leafburn_df <- boot_leafburn_df %>%
  mutate(fits=map(boot_samps,leafburn_boot_fit))
```

## Results

```{r,message=FALSE,warning=FALSE,echo=FALSE,fig.height=4}
boot_samp_leafburn <- function(i){
  boot_df <- sample_n(leafburn,nrow(leafburn),replace = TRUE)
}

leafburn_boots <- map(1:500,boot_samp_leafburn)

boot_leafburn_df <- tibble(boot_samps=leafburn_boots)

leafburn_boot_fit <- function(df){
  lm_fit <- lm(burntime~nitrogen+chlorine+potassium,data=df)
  return(tidy(lm_fit))
}

boot_leafburn_df <- boot_leafburn_df %>%
  mutate(fits=map(boot_samps,leafburn_boot_fit))

boot_leafburn_df %>% unnest(fits) %>%
  group_by(term) %>%
  ggplot(aes(x=estimate)) + geom_histogram() + 
  facet_wrap(~term)
```

## Conclusion

- We have introduced and demonstrated iteration in R using the `map` family of functions from the `purrr` package. 

- For more information, see [R for Data Science](https://r4ds.had.co.nz/), [Advanced R](https://adv-r.hadley.nz/), and [purrr Documentation](https://purrr.tidyverse.org/).

- A `purrr` cheatsheet is available [here](https://purrr.tidyverse.org/).

- The complete code for all examples is available [here](https://github.com/jmgraham30/SWBmodules).
